\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2025-06-25 doc/jss-accepted-version/tex/jss4956.tex
Trivial whitespace changes etc. in .tex output

% \usepackage{changes}
\usepackage[final]{changes}
\usepackage{changes}

@@

\Abstract{
Geostatistical spatial or spatiotemporal data are common across scientific fields.
% and the dimensions of these data continue to grow in space and time.
However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian Markov random fields (GMRFs), are computationally intensive and challenging for many users to implement.
Here, we introduce the \R\ package \sdmTMB, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent GMRFs using an SPDE-(stochastic partial differential equation) based approach.
SPDE matrices are constructed with \pkg{fmesher} and estimation is conducted via maximum marginal likelihood with \TMB\ or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
We describe the model and explore case studies that illustrate \sdmTMB's flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), hurdle models, cross-validation and anisotropy (directionally dependent spatial correlation).
% Additional functionality includes break-point effects, out-of-sample cross validation, and priors.
Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \sdmTMB\ can be an order of magnitude faster than \INLA.
We hope \sdmTMB \deleted{'s accessible interface} will help open this useful class of models to a wider field of geostatistical analysts.
}

@@

We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but a lack of evidence for spatial correlation after including a random field suggesting that residual spatial autocorrelation is alleviated by including the random field.
The specific p-value is dependent on the seed due to the randomization in the randomized quantile residuals.

@@

\deleted{
On inspecting the model output with \fct{summary}, we notice that the spatial random field SD in the positive model component has become very small once AR(1) spatiotemporal fields were added.
We therefore consider a model where these are turned off in the second component of the delta model and choose this model for further evaluation.
}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\textwidth]{Figures/timing-spatial-multipanel}
\caption[]{
Comparison of time to fit an SPDE spatial random field model with an intercept, one fixed-effect predictor, Gaussian error, and a sequence of SPDE resolutions to three dataset sizes.
Lines represent means and ribbons 90\% quantiles across 50 random iterations.
\pkg{VAST} and \pkg{tinyVAST} should be similar to \sdmTMB\ and so are not shown.
\pkg{inlabru} (version 2.10.1) used the empirical Bayes integration strategy and Gaussian approximation with \code{bru\_max\_iter = 1}, and the \fct{like} formulation.
\pkg{mgcv} (version 1.9.1) used \fct{bam}, \code{method = `fREML'}, and discretized covariates \citep{miller2019}.
Note that \pkg{spaMM} (version 4.4.16) only fits spatial, not spatiotemporal, models.
All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.
See Figure~\ref{fig:timing2} for a version with log-log axes.
}
\label{fig:timing}
\end{figure}

@@

\section{Speed testing related packages}\label{app:speed}
Here, we describe the methods underlying the speed testing in Figure~\ref{fig:timing}.
We generated a mesh that was consistent across simulated data sets for a given mesh resolution (Figure~\ref{fig:mesh-vis-timing}).
We did this by setting the \code{max.edge} argument, which controls the largest allowed triangle edge length.
We tested values of \code{max.edge} of 0.06, 0.075, 0.1, 0.15, and 0.2.
In Figure~\ref{fig:timing}, we report on the x-axis the number of mesh nodes (``vertices'') that result from each of these meshes.

>>>>

\usepackage{changes}

@@

\Abstract{
Geostatistical spatial or spatiotemporal data are common across scientific fields.
% and the dimensions of these data continue to grow in space and time.
However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian Markov random fields (GMRFs), are computationally intensive and challenging for many users to implement.
Here, we introduce the \R\ package \sdmTMB, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent GMRFs using an SPDE-(stochastic partial differential equation) based approach.
SPDE matrices are constructed with \pkg{fmesher} and estimation is conducted via maximum marginal likelihood with \TMB\ or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
We describe the model and explore case studies that illustrate \sdmTMB's flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), hurdle models, cross-validation and anisotropy (directionally dependent spatial correlation).
% Additional functionality includes break-point effects, out-of-sample cross validation, and priors.
Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \sdmTMB\ can be an order of magnitude faster than \INLA.
We hope \sdmTMB 's accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
}

@@

% We omit example code for Moran's I calculation for brevity but include it in the code supplement.
We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but a lack of evidence for spatial correlation after including a random field suggesting that residual spatial autocorrelation is alleviated by including the random field.
The specific p-value is dependent on the seed due to the randomization in the randomized quantile residuals.

@@

On inspecting the model output with \fct{summary}, we notice that the spatial random field SD in the positive model component has become very small once AR(1) spatiotemporal fields were added.
We therefore consider a model where these are turned off in the second component of the delta model and choose this model for further evaluation.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\textwidth]{Figures/timing-spatial-multipanel.pdf}
\caption[]{
Comparison of time to fit an SPDE spatial random field model with an intercept, one fixed-effect predictor, Gaussian error, and a sequence of SPDE resolutions to three dataset sizes.
Lines represent means and ribbons 90\% quantiles across 50 random iterations.
\pkg{VAST} and \pkg{tinyVAST} should be similar to \sdmTMB\ and so are not shown.
\pkg{inlabru} (version 2.10.1) used the empirical Bayes integration strategy and Gaussian approximation with \code{bru\_max\_iter = 1}, and the \fct{like} formulation.
\pkg{mgcv} (version 1.9.1) used \fct{bam}, \code{method = `fREML'}, and discretized covariates \citep{miller2019}.
Note that \pkg{spaMM} (version 4.4.16) only fits spatial, not spatiotemporal, models.
All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.
See Figure~\ref{fig:timing2} for a version with log-log axes.
}
\label{fig:timing}
\end{figure}

@@

\section{Speed testing related packages}\label{app:speed}

Here, we describe the methods underlying the speed testing in Figure~\ref{fig:timing}.
We generated a mesh that was consistent across simulated data sets for a given mesh resolution (Figure~\ref{fig:mesh-vis-timing}).
We did this by setting the \code{max.edge} argument, which controls the largest allowed triangle edge length.
We tested values of \code{max.edge} of 0.06, 0.075, 0.1, 0.15, and 0.2.
In Figure~\ref{fig:timing}, we report on the x-axis the number of mesh nodes (``vertices'') that result from each of these meshes.
